---
title: "Combine datasets for Switzerland"
author: "Pepa Aran"
date: "2023-07-24"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(dplyr)
library(stringr)
library(terra)
library(ggplot2)
library(tidyterra)
library(maps)
library(rnaturalearth)
library(rnaturalearthdata)
```

# Combining datasets for the swiss territory

Since the WorldClim data is too heavy to process locally, we will crop the global maps and work with the territory covering Switzerland only. 

## Subset data and compute VPD

We start by loading the WorldClim data and saving the cropped files.
```{r}
# Path for WorldClim data
path_worldclim <- "~/data/archive/worldclim_fick_2017/data"

# Read the temperature data for January
temp_worldclim <- terra::rast(paste0(path_worldclim, "/wc2.1_30s_tavg_01.tif"))

# Get the map data for Switzerland
switzerland_map <- ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf")

# Crop approximately the area of Switzerland
temp_worldclim <- terra::crop(x = temp_worldclim,
                              y = terra::vect(switzerland_map))

# Plot average January temperature in Switzerland
ggplot() +
  tidyterra::geom_spatraster(data = temp_worldclim) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

We can repeat the exercise for the WATCH-WFDEI data and appreciate the difference in resolution. 
```{r}
# Path for WATCH-WFDEI data (directory containing the .nc files)
path_watch <- "~/data/archive/wfdei_weedon_2014/data/Tair_daily"

# Get the list of .nc files in the directory
file_list <- list.files(path_watch, pattern = "*.nc", full.names = TRUE)

# Read the temperature data for January 1979, K -> C
temp_watch <- terra::rast(file_list[1]) - 273.15

# Crop approximately the area of Switzerland for all daily data
temp_watch <- terra::crop(x = temp_watch,
                              y = terra::vect(switzerland_map))

# Plot month average
ggplot() +
  tidyterra::geom_spatraster(data = temp_watch |>
                               terra::app(mean, na.rm = TRUE)) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

In the next code chunk, we read all WorldClim files, crop the data, compute VPD and save the resulting raster.
```{r}
# Define function to compute VPD
calc_vpd_from_vapr <- function(values){
  # Use the raster layers as a vector
    tc <- values[1]       # temperature in C
    vapr <- values[2]     # water vapor pressure in Pa
  
  # Calculate VPD
  611.0 * exp( (17.27 * tc)/(tc + 237.3) ) - vapr
}

process_worldclim <- function(
    data_path,   # path for WorldClim data
    month,       # month number in character, e.g. "01"
    crop_map,    # map from rnaturalearth to crop global map
    filename,    # name for the output file
    cores        # number of cores for parallel computation
){
  # Read temperature data and crop
  path_temp <- paste0(data_path, "/wc2.1_30s_tavg_", month, ".tif")
  temp_worldclim <- terra::rast(path_temp) |>
    terra::crop(vect(crop_map))
  
  # Read water vapor pressure data and crop
  path_vapr <- paste0(data_path, "/wc2.1_30s_vapr_", month, ".tif")
  vapr_worldclim <- terra::rast(path_vapr) |>
    terra::crop(vect(crop_map)) / 1000         # kPa -> Pa
  
  # Compute VPD from the WorldClim data
  # vpd_worldclim <- c(temp_worldclim, vapr_worldclim) |>
  #   terra::app(calc_vpd_from_vapr)
  vpd_worldclim <- c(temp_worldclim, vapr_worldclim) |>
    terra::app(fun = function(i, ff) ff(i),
               cores = cores,
               ff = calc_vpd_from_vapr)        # must export function to the nodes
  
  # Return a single raster with two layers
  terra::add(temp_worldclim) <- vpd_worldclim
  names(temp_worldclim) <- paste0(c("tavg_", "vpd_"), month)
  temp_worldclim |>
    terra::writeRaster(
      filename = filename,
      overwrite = TRUE
    )
}

# Process data for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  process_worldclim(
    data_path = "~/data/archive/worldclim_fick_2017/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/wc2.1_30s_CH_tavg_vpd_", m, ".tif"),
    cores = 1
  )
}
```

We repeat the exercise for all WATCH-WFDEI data. Although the resolution of the data is much coarser, the fact that we have daily values makes the processing quite time consuming. We will work with an example first and then optimize the runtime to downscale all data in 1979-2018.
```{r}
# Define functions to get global monthly climatology
# (1979-2000 to align with WorldClim)
source("R/derive_monthly_temperature.R")

# Get average temperature for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  derive_monthly_temperature(
    data_path = "~/data/archive/wfdei_weedon_2014/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/watch_CH_tavg_", m, ".tif"),
    cores = 1
  )
}


# Let's plot the August temperature as example
# plot_worldclim("data/watch_CH_tavg_08.tif", "tavg_08")
```

```{r}
# Define function to compute VPD
calc_vpd_from_qair <- function(tc, qair, patm){
  # Use the raster layers as a vector
  # tc:   temperature in C
  # qair: specific humidity in kg/kg
  # patm: surface  (atmospheric) pressure in Pa
  
  ## calculate the mass mixing ratio of water vapor to dry air (dimensionless)
  wair <- qair / (1 - qair)
  
  # Define constants
  # kR  = 8.3143   # universal gas constant, J/mol/K (Allen, 1973)
  # kMv = 18.02    # molecular weight of water vapor, g/mol (Tsilingiris, 2008)
  # kMa = 28.963   # molecular weight of dry air, g/mol (Tsilingiris, 2008)
  
  # Calculate water vapor pressure 
  rv <- 0.4613929  # rv <- kR / kMv
  rd <- 0.2870663  # rd <- kR / kMa
  vapr <- patm * wair * rv / (rd + wair * rv)
  
  # Calculate VPD
  611.0 * exp( (17.27 * tc)/(tc + 237.3) ) - vapr
  
  ## this empirical equation may lead to negative values for VPD
  ## (happens very rarely). assume positive...
  # max( 0.0, vpd )
}

# Define functions to get global monthly climatology
# (1979-2000 to align with WorldClim)
# Simultaneously save daily cropped rasters for temperature and VPD in data/
source("R/derive_monthly_vpd.R")

# Get average temperature for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  derive_monthly_vpd(
    data_path = "~/data/archive/wfdei_weedon_2014/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/watch_CH_vpd_", m, ".tif"),
    cores = 1
  )
}


# Let's plot the August temperature as example
# plot_worldclim("data/watch_CH_vpd_08.tif", "vpd_08")
```

### Plotting data subsets

Finally, let's plot the data. To make it easier, we will define a function that we can reuse which uses the data saved in files.
```{r}
# Function that plots data in file
plot_worldclim <- function(filename, layer_name){
  # Read file
  r <- terra::rast(filename)
  
  # Define plot legend and range
  if(grepl("tavg", layer_name)){
    ggplot() +
      tidyterra::geom_spatraster(data = r[[layer_name]] ) +
      scale_fill_viridis_c(
        na.value = NA,
        name = "Temperature (C) \n",
        limits = c(-40, 40)
        ) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        )
  }else if(grepl("vpd", layer_name)){
    ggplot() +
      tidyterra::geom_spatraster(data = r[[layer_name]] ) +
      scale_fill_viridis_c(
        na.value = NA,
        name = "Vapor Pressure Deficit (Pa) \n",
        limits = c(0, 3000)
        ) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        )
  }
  
}

plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_01.tif",
               layer_name = "tavg_01")
plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_01.tif",
               layer_name = "vpd_01")
plot_worldclim("data/watch_CH_tavg_01.tif",
               layer_name = "tavg_01")
plot_worldclim("data/watch_CH_vpd_01.tif",
               layer_name = "vpd_01")


plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_07.tif",
               layer_name = "tavg_07")
plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_07.tif",
               layer_name = "vpd_07")
```

#
