---
title: "Reading and processing Worldclim data"
author: "Pepa Aran"
date: "2023-07-07"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(dplyr)
library(stringr)
library(terra)
library(ggplot2)
library(tidyterra)
```

The 30sec gridded data from WorldClim can be directly downloaded from https://www.worldclim.org/data/worldclim21.html. We are interested in the following variables:

- **tavg**: average monthly temperature ($^\text{o}$C)
- **vapr**: water vapor pressure (kPa)
- **elev**: elevation (m)

Let's take a look at the average temperature across the globe in January, between the years 1970-2000.
```{r}
# Path for WorldClim data
path_worldclim <- "~/data/archive/worldclim_fick_2017/data"

# Read the temperature data for January
temp_worldclim <- terra::rast(paste0(path_worldclim, "/wc2.1_30s_tavg_01.tif"))

# Plot average
ggplot() +
  tidyterra::geom_spatraster(data = temp_worldclim) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

We can also observe the values for water vapor pressure, for the same time and locations. Since we will work with Pa, we need to do a unit transformation when reading the data.
```{r}
# Path for WorldClim data
path_worldclim <- "~/data/archive/worldclim_fick_2017/data"

# Read the water vapor pressure data for January, kPa -> Pa
vapr_worldclim <- terra::rast(paste0(path_worldclim, "/wc2.1_30s_vapr_01.tif")) * 1000

# Plot average
ggplot() +
  tidyterra::geom_spatraster(data = vapr_worldclim) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Water Vapor Pressure (Pa) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

From the water vapor pressure and temperature, we can easily compute our variable of interest VPD (vapor pressure deficit) as follows:

$$ VPD = 0.611 \exp \left({\frac{17.27 T}{T + 237.3}} \right) - P $$

where $T$ stands for temperature and $P$ for water vapor pressure.

```{r eval = FALSE}
# Define function to compute VPD
calc_vpd_from_vapr <- function(values){
  # Use the raster layers as a vector
    tc <- values[1]       # temperature in C
    vapr <- values[2]     # water vapor pressure in Pa
  
  # Calculate VPD
  611.0 * exp( (17.27 * tc)/(tc + 237.3) ) - vapr
}

# Compute VPD from the WorldClim data
# vpd_worlclim <- c(temp_worldclim, vapr_worldclim) |>
#   terra::app(calc_vpd_from_vapr)
vpd_worldclim <- c(temp_worldclim, vapr_worldclim) |>
  terra::app(fun = function(i, ff) ff(i),
             cores = 4,
             ff = calc_vpd_from_vapr,        # must export function to the nodes
             filename = paste0(here::here(), "/data/wc2.1_30s_vpd_01.tif"))

# Plot average
ggplot() +
  tidyterra::geom_spatraster(data = vpd_worldclim) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Vapor Pressure Deficit (Pa) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```


