---
title: "Reading and processing WATCH-WFDEI data"
author: "Pepa Aran"
date: "2023-07-07"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(dplyr)
library(stringr)
library(terra)
library(ggplot2)
library(tidyterra)
```

WATCH-WFDEI provides monthly climate fields at 0.5 degree resolution from 1979 to 2012. 
The data can be downloaded from the IIASA website via ftp for each variable separately:
- **Tair** is the daily average (of 3-hourly) near surface air temperature (in Kelvin) at 2m at time stamp.
- **Qair** is the daily average specific humidity (in kg/kg) at 2m above surface
- **PSurf** is the daily average surface pressure (in hPa) 

After downloading the data from the IIASA website via ftp, we can read the netCDF files and extract the information we're interested in. For example, `Tair_daily_WFDEI_197901.nc` contains average temperature values for each day of January 1979 and each pixel in a 0.5 degree grid covering the whole globe. Therefore, to get the time series of temperature for a given pixel

Let's start by reading the data for the month of January 1979 only.
```{r}
# Path for WATCH-WFDEI data (directory containing the .nc files)
path_watch <- "~/data/archive/wfdei_weedon_2014/data/Tair_daily"

# Get the list of .nc files in the directory
file_list <- list.files(path_watch, pattern = "*.nc", full.names = TRUE)

# Read the temperature data for January, K -> C
temp_watch <- terra::rast(file_list[1]) - 273.15

# Plot month average
ggplot() +
  tidyterra::geom_spatraster(data = temp_watch |>
                               terra::app(mean, na.rm = TRUE)) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

Now let's read the specific humidity and surface pressure data, as we did for temperature.
```{r}
# Path for WATCH-WFDEI data (directory containing the .nc files)
path_watch_qair <- "~/data/archive/wfdei_weedon_2014/data/Qair_daily"
path_watch_psurf <- "~/data/archive/wfdei_weedon_2014/data/PSurf_daily"

# Get the list of .nc files in the directory
file_list_qair <- list.files(path_watch_qair, pattern = "*.nc", full.names = TRUE)
file_list_psurf <- list.files(path_watch_psurf, pattern = "*.nc", full.names = TRUE)

# Read the data for January 1979
qair_watch <- terra::rast(file_list_qair[1])
psurf_watch <- terra::rast(file_list_psurf[1]) * 100   # hPa -> Pa

# Plot month average
ggplot() +
  tidyterra::geom_spatraster(data = qair_watch |>
                               terra::app(mean, na.rm = TRUE)) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Specific humidity (kg/kg) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )

ggplot() +
  tidyterra::geom_spatraster(data = psurf_watch |>
                               terra::app(mean, na.rm = TRUE)) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Surface pressure (Pa) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

As we did for the WorldClim data, we can compute VPD (vapor pressure deficit) from the specific humidity, elevation and temperature data provided in WATCH-WFDEI:

$$ VPD = 0.611 \exp \left({\frac{17.27 T}{T + 237.3}} \right) - P $$

where $T$ stands for temperature and $P$ for water vapor pressure.

```{r}
# Define function to compute VPD
calc_vpd_from_qair <- function(values){
  # Use the raster layers as a vector
    tc <- values[1]       # temperature in C
    qair <- values[2]     # specific humidity in kg/kg
    patm <- values[3]     # surface  (atmospheric) pressure in Pa
  
  ## calculate the mass mixing ratio of water vapor to dry air (dimensionless)
    wair <- qair / (1 - qair)
    
  # Define constants
    # kR  = 8.3143   # universal gas constant, J/mol/K (Allen, 1973)
    # kMv = 18.02    # molecular weight of water vapor, g/mol (Tsilingiris, 2008)
    # kMa = 28.963   # molecular weight of dry air, g/mol (Tsilingiris, 2008)
    
  # Calculate water vapor pressure 
    rv <- 0.4613929  # rv <- kR / kMv
    rd <- 0.2870663  # rd <- kR / kMa
    vapr = patm * wair * rv / (rd + wair * rv)
  
  # Calculate VPD
  611.0 * exp( (17.27 * tc)/(tc + 237.3) ) - vapr
  
  ## this empirical equation may lead to negative values for VPD
  ## (happens very rarely). assume positive...
  # max( 0.0, vpd )
}

# Compute VPD from the WorldClim data
vpd_watch <- c(temp_watch, qair_watch, psurf_watch) |>
  terra::app(calc_vpd_from_qair)
# vpd_watch <- c(temp_watch, qair_watch, psurf_watch) |>
#   terra::app(fun = function(i, ff) ff(i),
#              cores = 4,
#              ff = calc_vpd_from_qair,        # must export function to the nodes
#              filename = paste0(here::here(), "/data/vpd_daily_WFDEI_197901.tif"))

# Plot average
ggplot() +
  tidyterra::geom_spatraster(data = vpd_watch) +
  scale_fill_viridis_c(
    na.value = NA,
    name = " Vapor Pressure Deficit (Pa) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

```{r}
# Crop VPD raster to Switzerland only and plot
switzerland_map <- rnaturalearth::ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf")

ggplot() +
  tidyterra::geom_spatraster(data = vpd_watch |>
                               terra::crop(vect(switzerland_map))) +
  scale_fill_viridis_c(
    na.value = NA,
    name = " Vapor Pressure Deficit (Pa) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

