---
title: "Combine datasets for Switzerland"
author: "Pepa Aran"
date: "2023-07-24"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(dplyr)
library(stringr)
library(terra)
library(ggplot2)
library(tidyterra)
library(maps)
library(rnaturalearth)
library(rnaturalearthdata)
```

# Combining datasets for the swiss territory

Since the WorldClim data is too heavy to process locally, we will crop the global maps and work with the territory covering Switzerland only. 

## Subset data and compute VPD

We start by loading the WorldClim data and saving the cropped files.
```{r}
# Path for WorldClim data
path_worldclim <- "~/data/archive/worldclim_fick_2017/data"

# Read the temperature data for January
temp_worldclim <- terra::rast(paste0(path_worldclim, "/wc2.1_30s_tavg_01.tif"))

# Get the map data for Switzerland
switzerland_map <- ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf")

# Crop approximately the area of Switzerland
temp_worldclim <- terra::crop(x = temp_worldclim,
                              y = terra::vect(switzerland_map))

# Plot average January temperature in Switzerland
ggplot() +
  tidyterra::geom_spatraster(data = temp_worldclim) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

We can repeat the exercise for the WATCH-WFDEI data and appreciate the difference in resolution. 
```{r}
# Path for WATCH-WFDEI data (directory containing the .nc files)
path_watch <- "~/data/archive/wfdei_weedon_2014/data/Tair_daily"

# Get the list of .nc files in the directory
file_list <- list.files(path_watch, pattern = "*.nc", full.names = TRUE)

# Read the temperature data for January 1979, K -> C
temp_watch <- terra::rast(file_list[1]) - 273.15

# Crop approximately the area of Switzerland for all daily data
temp_watch <- terra::crop(x = temp_watch,
                              y = terra::vect(switzerland_map))

# Plot month average
ggplot() +
  tidyterra::geom_spatraster(data = temp_watch |>
                               terra::app(mean, na.rm = TRUE)) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Temperature (C) \n"
    ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    )
```

In the next code chunk, we read all WorldClim files, crop the data, compute VPD and save the resulting raster.
```{r}
# Define function to compute VPD
source("R/calc_vpd_from_vapr.R")

process_worldclim <- function(
    data_path,   # path for WorldClim data
    month,       # month number in character, e.g. "01"
    crop_map,    # map from rnaturalearth to crop global map
    filename,    # name for the output file
    cores        # number of cores for parallel computation
){browser()
  # Read temperature data and crop
  path_tavg <- paste0(data_path, "/wc2.1_30s_tavg_", month, ".tif")
  tavg <- terra::rast(path_tavg) |>
    terra::crop(terra::vect(crop_map))
  
  # Read water vapor pressure data and crop
  path_vapr <- paste0(data_path, "/wc2.1_30s_vapr_", month, ".tif")
  vapr <- terra::rast(path_vapr) |>
    terra::crop(vect(crop_map)) * 1000         # kPa -> Pa
  
  # Compute VPD from the WorldClim data
  vpd <- calc_vpd_from_vapr(tavg, vapr)
  # vpd_worldclim <- c(temp_worldclim, vapr_worldclim) |>
  #   terra::app(fun = function(i, ff) ff(i),
  #              cores = cores,
  #              ff = calc_vpd_from_vapr)   # must export function to the nodes
  
  # Return a single raster with two layers
  terra::add(temp_worldclim) <- vpd
  names(vpd) <- paste0("vpd_", month)
  temp_worldclim |>
    terra::writeRaster(
      filename = filename,
      overwrite = TRUE
    )
}

# Process data for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  process_worldclim(
    data_path = "~/data/archive/worldclim_fick_2017/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/CH_vpd_", m, ".tif"),
    cores = 1
  )
}
```

We repeat the exercise for all WATCH-WFDEI data. Although the resolution of the data is much coarser, the fact that we have daily values makes the processing quite time consuming. We will work with an example first and then optimize the runtime to downscale all data in 1979-2018.
```{r}
# Define functions to get global monthly climatology
# (1979-2000 to align with WorldClim)
source("R/derive_monthly_temperature.R")

# Get average temperature for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  derive_monthly_temperature(
    data_path = "~/data/archive/wfdei_weedon_2014/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/watch_CH_tavg_", m, ".tif"),
    cores = 1
  )
}


# Let's plot the August temperature as example
# plot_worldclim("data/watch_CH_tavg_08.tif", "tavg_08")
```

```{r}
# Define function to compute VPD
source("R/calc_vpd_from_qair.R")

# Define functions to get global monthly climatology
# (1979-2000 to align with WorldClim)
# Simultaneously save daily cropped rasters for temperature and VPD in data/
source("R/derive_monthly_vpd.R")

# Get average temperature for all 12 months
for(m in c(paste0("0", 1:9), "10", "11", "12")){
  derive_monthly_vpd(
    data_path = "~/data/archive/wfdei_weedon_2014/data",
    month = m,
    crop_map = ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf"),
    filename = paste0("data/watch_CH_vpd_", m, ".tif"),
    cores = 1
  )
}


# Let's plot the August temperature as example
# plot_worldclim("data/watch_CH_vpd_08.tif", "vpd_08")
```

### Plotting data subsets

Finally, let's plot the data. To make it easier, we will define a function that we can reuse which uses the data saved in files.
```{r}
# Function that plots data in file
plot_worldclim <- function(filename, layer_name){
  # Read file
  r <- terra::rast(filename)
  
  # Define plot legend and range
  if(grepl("tavg", layer_name)){
    ggplot() +
      tidyterra::geom_spatraster(data = r[[layer_name]] ) +
      scale_fill_viridis_c(
        na.value = NA,
        name = "Temperature (C) \n",
        limits = c(-40, 40)
        ) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        )
  }else if(grepl("vpd", layer_name)){
    ggplot() +
      tidyterra::geom_spatraster(data = r[[layer_name]] ) +
      scale_fill_viridis_c(
        na.value = NA,
        name = "Vapor Pressure Deficit (Pa) \n",
        limits = c(0, 3000)
        ) +
      theme_bw() +
      theme(
        legend.position = "bottom"
        )
  }
  
}

plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_01.tif",
               layer_name = "tavg_01")
plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_08.tif",
               layer_name = "vpd_08")
plot_worldclim("data/watch_CH_tavg_01.tif",
               layer_name = "tavg_01")
plot_worldclim("data/watch_CH_vpd_08.tif",
               layer_name = "vpd_08")


plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_07.tif",
               layer_name = "tavg_07")
plot_worldclim("data/wc2.1_30s_CH_tavg_vpd_07.tif",
               layer_name = "vpd_07")
```

# Scratch paper to compare VPD calculations
```{r}
# Check VPD calculation
source("R/calc_vpd_from_vapr.R")
source("R/calc_vpd_from_qair.R")

# Monthly averages are much larger for WATCH-WFDEI than for WorldClim

data_path <- "~/data/archive/wfdei_weedon_2014/data"
month <- "07"
crop_map <- ne_countries(scale = "medium", country = "Switzerland", returnclass = "sf")

# Read data from 1979 to 2000
files_tair <- list.files(paste0(data_path, "/Tair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[1:22]
files_qair <- list.files(paste0(data_path, "/Qair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[1:22]
files_psurf <- list.files(paste0(data_path, "/PSurf_daily"),
                          pattern = paste0("*", month,".nc"),
                          full.names = TRUE)[1:22]

y <- 1 # 1979
tair <- terra::rast(files_tair[y]) |>
  terra::crop(terra::vect(crop_map)) - 273.15   # K -> C
qair <- terra::rast(files_qair[y]) |>
  terra::crop(terra::vect(crop_map)) 
psurf <- terra::rast(files_psurf[y]) |>
  terra::crop(terra::vect(crop_map))

# Compute VPD
vpd <- calc_vpd_from_qair(t = tair,
                          q = qair,
                          P = psurf)
names(vpd) <- gsub("Tair", "vpd", names(vpd))

ggplot() +
  tidyterra::geom_spatraster(data = vpd[[1]] ) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Vapor Pressure Deficit (Pa) \n",
    limits = c(0, 1000)
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  )


# Now for Worldclim
data_path <- "~/data/archive/worldclim_fick_2017/data"

path_tavg <- paste0(data_path, "/wc2.1_30s_tavg_", month, ".tif")
tavg <- terra::rast(path_tavg) |>
  terra::crop(terra::vect(crop_map))

# Read water vapor pressure data and crop
path_vapr <- paste0(data_path, "/wc2.1_30s_vapr_", month, ".tif")
vapr <- terra::rast(path_vapr) |>
  terra::crop(vect(crop_map)) * 1000         # kPa -> Pa

# Compute VPD from the WorldClim data
vpd <- calc_vpd_from_vapr(t = tavg, 
                          e = vapr)
names(vpd) <- paste0("vpd_", month)

ggplot() +
  tidyterra::geom_spatraster(data = vpd ) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "Vapor Pressure Deficit (Pa) \n",
    limits = c(0, 1000)
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  )
```



# Function to plot data
```{r}
plot_temp <- function(r){
  ggplot() +
    tidyterra::geom_spatraster(data = r) +
    scale_fill_viridis_c(
      na.value = NA,
      name = "Temperature (C) \n",
      limits = c(-40, 40)
      ) +
    theme_bw() +
    theme(
      legend.position = "bottom"
      )
}

plot_vpd <- function(r){
  ggplot() +
    tidyterra::geom_spatraster(data = r) +
    scale_fill_viridis_c(
      na.value = NA,
      name = "Vapor Pressure Deficit (Pa) \n",
      limits = c(0, 1000)
      ) +
    theme_bw() +
    theme(
      legend.position = "bottom"
      )
}
```






profvis({
lapply(seq_along(files_daily),
      FUN = function(y){
        # Resample daily data with bilinear interpolation to 1/12 deg
        daily <- terra::rast(files_daily[y]) |>
          terra::resample(clim_monthly_worldclim) |>
          terra::crop(clim_monthly_worldclim)
        
        # De-bias
        daily <- terra::app(c(clim_monthly_worldclim, daily),
                             fun = function(i, ff) ff(i),
                             cores = cores,
                             ff = function(r){
                               n <- (length(r) - 2)/2   # number of days
                               t <- r[2 + (1:n) ] + r[1]
                               vpd <- r[2 + n + (1:n) ] + r[2]
                               
                               # Return temp and vpd
                               c(t, vpd)
                             })
        
        # Save in file
        terra::writeRaster(daily, 
                           filename = paste0("data/tair_vpd_daily_downscaled_",
                                             1990+y, month, ".tif"),
                           overwrite = TRUE)
      })
})
