---
title: "Downscaling workflow"
author: "Pepa Aran"
date: "2023-07-27"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(dplyr)
library(stringr)
library(terra)
library(ggplot2)
library(tidyterra)
library(maps)
library(rnaturalearth)
library(rnaturalearthdata)
```

# Downscaling temperature and VPD

Here we present the full workflow for the project, from raw rater data to the downscaled daily values. This code is then moved to a function `downscale_temp_vpd()` in order to run the workflow for all months of the year.

Firstly, let's define the raw data paths and the map of Switzerland. We crop the original data to a smaller geographical region because of the high computation costs that global maps require.
```{r}
# Path for WorldClim data
path_worldclim <- "~/data/archive/worldclim_fick_2017/data"

# Path for WATCH-WFDEI data (subdirectory contains the .nc files)
path_watch <- "~/data/archive/wfdei_weedon_2014/data"

# Create Switzerland map object
crop_map <- ne_countries(scale = "medium", 
                          country = "Switzerland", 
                          returnclass = "sf")

# Load functions to compute VPD
source("R/calc_vpd_from_qair.R")
source("R/calc_vpd_from_vapr.R")
```

The main parts of the downscaling workflow are the following:

- Calculating VPD from the raw WATCH-WFDEI data (save daily values in files).
- Calculate average monthly climatology from the WATCH-WFDEI daily data.
- Read WorldClim data and calculate VPD.
- Resample WATCH-WFDEI-derived climatology to WorldClim resolution and compute the climatology bias (save bias in files). $VPD_{bias} = VPD_{WATCH} - VPD_{WorldClim}$
- Resample daily data and substract the bias. $VPD_t = VPD_{WATCH, \; t} - VPD_{bias}$

```{r}
# Define function to get global monthly climatology
# (1979-2000 to align with WorldClim)

# Work with data from January only
month <- "01"

# Read WATCH-WFDEI data from 1979 to 2000
files_tair <- list.files(paste0(path_watch, "/Tair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[1:22]
files_qair <- list.files(paste0(path_watch, "/Qair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[1:22]
files_psurf <- list.files(paste0(path_watch, "/PSurf_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[1:22]

# Compute average monthly climatology, parallelized
cores <- 1
clim_monthly <- lapply(1:22,
                       FUN = function(y){
                         # Read a year's data and crop
                         tair <- terra::rast(files_tair[y]) |>
                           terra::crop(terra::vect(crop_map)) - 273.15   # K -> C
                         qair <- terra::rast(files_qair[y]) |>
                           terra::crop(terra::vect(crop_map)) 
                         psurf <- terra::rast(files_psurf[y]) |>
                           terra::crop(terra::vect(crop_map))
                         
                         # Compute VPD
                         vpd <- calc_vpd_from_qair(t = tair,
                                                   q = qair,
                                                   P = psurf)
                         names(vpd) <- gsub("Tair", "vpd", names(vpd))
                         names(tair) <- gsub("Tair", "tavg", names(tair))
                         
                         # Save daily VPD and temperature calculations at 0.5deg
                         terra::writeRaster(c(tair, vpd), 
                                            filename = paste0("data/tair_vpd_daily_",
                                                              1978+y, month, ".tif"),
                                            overwrite = TRUE)
                         
                         # Compute monthly average
                         tair_monthly <- tair |>
                           terra::app(mean, na.rm = TRUE)
                         names(tair_monthly) <- paste0("tair_", month)
                         
                         vpd_monthly <- vpd |>
                           terra::app(mean, na.rm = TRUE)
                         names(vpd_monthly) <- paste0("vpd_", month)


                         c(tair_monthly, vpd_monthly)
                       }) |>
    terra::rast() |>                      # convert to single rast
    terra::app(fun = function(i, ff) ff(i),
               cores = cores,
               ff = function(r){
                 # Separate temperature and vpd
                 mean_t <- mean(r[seq(from = 1, to = length(r), by = 2)],
                                na.rm = TRUE)
                 mean_vpd <- mean(r[seq(from = 2, to = length(r), by = 2)],
                                  na.rm = TRUE)
                 c(mean_t, mean_vpd)
               })                         # avg over 22 years
# Match names to WorldClim data
names(clim_monthly) <- c(paste0("tavg_", month),
                         paste0("vpd_", month))

# Read WorldClim data
# Read temperature data and crop
file_tavg <- paste0(path_worldclim, "/wc2.1_30s_tavg_", month, ".tif")
tavg <- terra::rast(file_tavg) |>
  terra::crop(terra::vect(crop_map))

# Read water vapor pressure data and crop
file_vapr <- paste0(path_worldclim, "/wc2.1_30s_vapr_", month, ".tif")
vapr <- terra::rast(file_vapr) |>
  terra::crop(vect(crop_map)) * 1000         # kPa -> Pa
  
vpd <- calc_vpd_from_vapr(t = tavg, 
                          e = vapr)
names(vpd) <- paste0("vpd_", month)
clim_monthly_worldclim <- c(tavg, vpd)
```


Now that we have computed the monthly climatology from WATCH-WFDEI and loaded the WorldClim average monthly climatology data, we can move on to compute the bias for the grid with finer spatial resolution.
```{r}
# Let's resample the spatial grid of the daily data (1/2 deg) to match
# the finer WorldClim grid (1/12 deg)

# Resampling computed with bilinear interpolation 
clim_monthly_resampled <- terra::resample(
  clim_monthly,            # SpatRaster to be resampled
  clim_monthly_worldclim)  # SpatRaster with the goal geometry

clim_monthly_resampled <- terra::crop(clim_monthly_resampled,
                                           clim_monthly_worldclim)

# Calculate climatic bias
clim_monthly_bias <- clim_monthly_resampled - clim_monthly_worldclim


```

Finally, we can de-bias the original daily temperature and vpd using the long-term climatologic trends for the finer grid. We save those results in the data/ folder.
```{r}
# Calculate VPD for Switzerland after year 2000
# Read WATCH-WFDEI data from 2000 to 2018
files_tair <- list.files(paste0(path_watch, "/Tair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[23:40]
files_qair <- list.files(paste0(path_watch, "/Qair_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[23:40]
files_psurf <- list.files(paste0(path_watch, "/PSurf_daily"),
                         pattern = paste0("*", month,".nc"),
                         full.names = TRUE)[23:40]

# Compute VPD and temperature for the remaining years of daily data
cores <- 4
lapply(seq_along(files_tair),
       FUN = function(y){
         # Read a year's data and crop
         tair <- terra::rast(files_tair[y]) |>
           terra::crop(terra::vect(crop_map)) - 273.15   # K -> C
         qair <- terra::rast(files_qair[y]) |>
           terra::crop(terra::vect(crop_map)) 
         psurf <- terra::rast(files_psurf[y]) |>
           terra::crop(terra::vect(crop_map))
         
         # Compute VPD
         vpd <- calc_vpd_from_qair(t = tair,
                                   q = qair,
                                   P = psurf)
         names(vpd) <- gsub("Tair", "vpd", names(vpd))
         names(tair) <- gsub("Tair", "tavg", names(tair))
         
         # Save daily VPD and temperature calculations at 0.5deg
         terra::writeRaster(c(tair, vpd), 
                            filename = paste0("data/tair_vpd_daily_",
                                              2000+y, month, ".tif"),
                            overwrite = TRUE)
       })

# Read and resample daily data
files_daily <- paste0("data/tair_vpd_daily_", 1979:2018, month, ".tif") 

# Define a simple function to de-bias
debias <- function(r, n){
  # n: number of days in month
  # r: raster including 
  #    t_bias | vpd_bias | t_01 ... t_n | vpd_01 ... vpd_n
  
  t <- r[2 + (1:n) ] + r[1]
  vpd <- r[2 + n + (1:n) ] + r[2]
  
  # Return temp and vpd
  c(t, vpd)
}

lapply(seq_along(files_daily),
       FUN = function(y){
         # Resample daily data with bilinear interpolation to 1/12 deg
         daily <- terra::rast(files_daily[y]) |>
           terra::resample(clim_monthly_worldclim) |>
           terra::crop(clim_monthly_worldclim)
         
         # Get number of days in month
         n <- terra::nlyr(daily)/2
         
         # De-bias
         daily <- terra::app(c(clim_monthly_worldclim, daily),
                              fun = function(i, ff) ff(i),
                              cores = cores,
                              ff = function(r) debias(r, n))
         
         # Save in file
         terra::writeRaster(daily, 
                            filename = paste0("data/tair_vpd_daily_downscaled_",
                                              1978+y, month, ".tif"),
                            overwrite = TRUE)
       })

```


